One-stage point transect distance sampling using iterated integrated nested Laplace approximations

### 2024-10-31 paper.tex
more edits

\subsection{Writing the model as a modified Poisson likelihood}

@@

The half-normal detection function depends on the strictly positive parameter $\sigma^2$. Since it is not possible to restrict the range of parameters to be strictly positive within \texttt{R-INLA} this model cannot be fitted using the standard INLA methodology.  Instead, we choose a log link and consider $g(\bs | \phi)$, where $\log\sigma^2 = \phi$. This is a non-linear function of $\phi$ which cannot be fitted using standard GLM methods.  To address this we use a novel approximate-method based on an iterative model fitting scheme. This method is implemented in the \texttt{R} package \texttt{inlabru} which is a wrapper for and an extension of the \texttt{R-INLA} package \citep{lindgren_inlabru_2024}.  Full details of the method can be found in the software paper, here we present a brief overview.

Let $\eta(\bm{u}) = \log \tilde{\lambda}(\bm{u})$ where $\bm{u}$ is the parameter vector of the predictor (which includes $\phi$ as well as any parameters for fixed and random effects).  Denote by $\bar{\eta}_{\bm{u}_0}$ the first-order Taylor approximation of $\eta(\bm{u})$ at the point $\bm{u}_0$.  That is
\begin{equation*}
\bar{\eta}_{\bm{u}_0} (\bm{u}) = \eta(\bm{u_0}) + \bm{B}_{\bm{u}_0}(\bm{u} - \bm{u}_0),
\end{equation*}
where $\bm{B}_{\bm{u}_0}$ is the derivative matrix of $\eta(\bm{u})$ evaluated at $\bm{u}_0$.  Since $\bar{\eta}_{\bm{u}_0}$ is linear in $\bm{u}$ it is possible to fit an approximate model in INLA by replacing the non-linear $\eta$ with the linear Taylor approximation $\bar{\eta}_{\bm{u_0}}$.

The question then is how to choose the linearisation point $\bm{u}_0$.  To find an `optimal' linearisation point we use a fixed point iteration scheme.  There are various possible choices to define an `optimal' choice and this affects the properties of the approximate inference method.  Here we describe the criteria used when fitting the model that produces the results presented in this paper.  However, more recently, other options have been added to \texttt{inlabru}, the software package that implements these methods.

Let $\bar{p}_{\bm{v}}$ denote the INLA approximate posterior distribution of the model parameters given the linearised model configuration at linearisation point $\bm{v}$.  New linearisation points are found by via the functional
\begin{equation}
	f(\ol{p}_{\bm{v}}) = (\wh{\bm{\theta}}_{\bm{v}},\wh{\bm{u}}_{\bm{v}})
\end{equation}
where
\begin{equation}
\wh{\bm{\theta}}_{\bm{v}} = \argmax_{\bm{\theta}} \ol{p}_{\bm{v}} ( \bm{\theta} | \bm{y}),
\end{equation}
the posterior mode for $\bm{\theta}$, and
\begin{equation}
\wh{\bm{u}}_{\bm{v}} = \argmax_{\bm{u}} \ol{p}_{\bm{v}} (\bm{u} | \bm{y}, \wh{\bm{\theta}}_{\bm{v}}),
\end{equation}
the joint conditional posterior mode for $\bm{u}$.  This functional defines a new linearisation point as the posterior mode for $\bm{\theta}$ under the linearised model configuration and the conditional mode for $\bm{u}$ given $\bm{\theta}$.  These posteriors are computed automatically during the model fitting using the standard INLA method and so this functional can be evaluated at no extra cost for each fit of the linearised model.

>>>>

\subsection{The modified Poisson likelihood}
\label{sec-lgcp}

@@

The half-normal detection function depends on the strictly positive parameter $\sigma^2$ and so this model cannot be fitted using the standard INLA methodology.  We choose a log link $\log\sigma^2 = \phi$ and use a novel approximate-method based on an iterative model fitting scheme. This method is implemented in the \texttt{R} package \texttt{inlabru} which is a wrapper for and an extension of the \texttt{R-INLA} package \citep{lindgren_inlabru_2024}. We briefly summarise the approach here, full details of the method can be found in \cite{lindgren_inlabru_2024}.

The family of models that can be fitted using the INLA methodology are known as latent Gaussian models.  These depend on a latent Gaussian vector of parameters which we denote $\bm{u}$. Parameters are related to observations via a predictor $\eta(\bm{u})$ which, for the LGCP model, is the log-intensity evaluated at the observed point locations and the integration scheme locations as described in Section \ref{sec-lgcp}.  Since $\phi$ is an element of $\bm{u}$, $\eta(\bm{u})$ is a non-linear function with respect to $\bm{u}$.

To motivate the approach, we consider a linearised approximation of $\eta(\bm{u})$ around point $\bm{u}_0$:
\begin{equation*}
\bar{\eta}_{\bm{u}_0} (\bm{u}) = \eta(\bm{u_0}) + \bm{B}_{\bm{u}_0}(\bm{u} - \bm{u}_0),
\end{equation*}
where $\bm{B}_{\bm{u}_0}$ is the derivative matrix of $\eta(\bm{u})$ evaluated at $\bm{u}_0$.  Since $\bar{\eta}_{\bm{u}_0}$ is linear in $\bm{u}$ it is possible to fit an approximate model in INLA by replacing the non-linear $\eta$ with the first-order Taylor approximation $\bar{\eta}_{\bm{u_0}}$.  This model configuration has $\bm{B}_{\bm{u}_0}$ as the model design matrix and $\eta(\bm{u}_0} - \bm{B}_{\bm{u}_0}\bm{u}_0$ as a known offset term.

The `optimal' linearisation point is defined via a fixed point iteration scheme. Let $\bar{p}_{\bm{v}}$ denote the INLA approximate posterior distribution of the model parameters given the linearised model configuration at linearisation point $\bm{v}$.  A new linearisation point is found by via the functional
\begin{equation}
	f(\ol{p}_{\bm{v}}) = (\wh{\bm{\theta}}_{\bm{v}},\wh{\bm{u}}_{\bm{v}})
\end{equation}
where
\begin{equation}
\wh{\bm{\theta}}_{\bm{v}} = \argmax_{\bm{\theta}} \ol{p}_{\bm{v}} ( \bm{\theta} | \bm{y}),
\end{equation}
the posterior mode for $\bm{\theta}$, and
\begin{equation}
\wh{\bm{u}}_{\bm{v}} = \argmax_{\bm{u}} \ol{p}_{\bm{v}} (\bm{u} | \bm{y}, \wh{\bm{\theta}}_{\bm{v}}),
\end{equation}
the joint conditional posterior mode for $\bm{u}$.  In other words, this functional defines a new linearisation point as the posterior mode for $\bm{\theta}$ under the linearised model configuration and the conditional mode for $\bm{u}$ given $\bm{\theta}$.  These posteriors are computed automatically during the model fitting using the standard INLA method and so this functional can be evaluated at no extra cost.
